---
title: "ggplot2"
author: "K. Shoemaker"
date: "DATA 2401"
output:
  slidy_presentation: default
  ioslides_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

```{r}

install.packages("ggthemes")

```

## `ggplot2`

We'll talk next about how to use a language for plotting, `ggplot2`

First up, install the package:

```{r, eval = F}
#install.packages("ggplot2", dependencies = T)
```

## Base vs `ggplot2`

Why `ggplot2` instead of the "base" graphics?

-   Base graphics are kinda clunky. It's harder work to get them looking nice.
-   The learning curve is steep, but worth it in the long run.
-   Similar to dplyr, it has a language, and that language is dataframe friendly.
-   Publication quality figures with a little work.

## Examples!

Iris scatter plot, with 2D density lines:

```{r, echo = F}
library(tidyverse)

iris %>% ggplot( aes(Sepal.Length, Sepal.Width, color = Species, shape = Species)) +
  geom_point() +
  geom_density2d() +
  ggtitle('IRIS') +
  theme_light()
```

------------------------------------------------------------------------

Faceted smoothing of the Iris data:

```{r, echo = FALSE}
iris %>%
  mutate(Species = 'ALL') %>%
  bind_rows(iris) %>%
  ggplot(aes(Petal.Length, Petal.Width, color = Species)) +
  geom_point() +
  geom_smooth(formula = y~x,method = "loess") +
  xlab('Petal Length') +
  ylab('Petal Width') +
  facet_wrap(~Species, scales = 'free') +
  theme_bw()
```

------------------------------------------------------------------------

"Tufte" style (the famous old "chart junk" graphics guy)

```{r, echo = F}
mtcars %>%
  rownames_to_column() %>%
  mutate(rowname = forcats::fct_reorder(rowname, mpg)) %>%
  ggplot(aes(rowname, mpg, label = rowname)) +
  geom_point() +
  geom_text(nudge_y = .3, hjust = 'left') +
  coord_flip() +
  ylab('Miles per gallon fuel consumption') +
  ylim(10, 40) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0, size = 16),
        axis.title.x = element_text(face = 'bold'),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.line.y = element_blank())
```

------------------------------------------------------------------------

A bubble plot with `mtcars`:

```{r, echo = F}
mtcars %>%
  ggplot(aes(mpg, qsec, size = disp, color = as.factor(am))) +
  geom_point() +
  scale_colour_discrete(name  ="Gear",
                        breaks=c(0, 1),
                        labels=c("Manual", "Automatic")) +
  scale_size_continuous(name = 'Displacement') +
  xlab('Miles per gallon') +
  ylab('1/4 mile time') +
  theme_light()
```

------------------------------------------------------------------------

A polar barplot of the mean `diamond` price per cut AND color.

```{r, echo = F}
diamonds2plot <- diamonds %>%
  group_by(cut, color) %>%
  summarise(price = mean(price)) %>%
  arrange(color, price) %>%
  ungroup() %>%
  mutate(id = row_number(),
         angle = 90 - 360 * (id - 0.5) / n())

diamonds2plot  %>%
  ggplot(aes(factor(id), price, fill = color, group = cut, label = cut)) +
  geom_bar(stat = 'identity', position = 'dodge') +
  geom_text(hjust = 0, angle = diamonds2plot$angle, alpha = .5) +
  coord_polar() +
  ggtitle('Mean dimond price') +
  ylim(-3000, 7000) +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = 'bold'))
```

------------------------------------------------------------------------

The economist theme:

```{r, echo = F}
economics %>%
  ggplot(aes(date, uempmed)) +
  geom_line() +
  expand_limits(y  = 0) +
  ggtitle('Median duration of unemployment [weeks]') +
  ylab('Median duration of unemployment [weeks]') +
  ggthemes::theme_economist_white() +
  theme(axis.title.x = element_blank())
```

## The Basics

-   **ggplot** - the main function where you specify the dataframe and the variables to plot
-   **geoms** - geometric objects that specify how the plot will be
    -   geom_bar(), geom_point(), geom_bar(), geom_line(), many *many* more.
-   **aes** - the "aesthetics" of how it will look
    -   shape, transparency, color, fill, line type.
-   **scales** - how the data will be plotted.
    -   continuous, discrete, log, etc.

## What that looks like

```{r, eval = F}
ggplot(data = dataset, mapping = aes(x = x_variable, y = y_variable)) +
   geom_xxx() +
   other options
```

-   Specify your data and variables inside the `ggplot` function.
-   Everything else gets layered on top of this, such as the geometric objects or statistical models

## Your First `ggplot`

The `iris` dataset:

```{r}
library(tidyverse)
head(iris)
```

## The Plot!

```{r, fig.height = 4}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) + geom_point()
```

Note: this is the basis of the first plot in the slides!

## Customization

`ggplots` are almost infinitely customizeable:

-   Titles/labels
-   colors
-   shapes

For more information than I could ever give you:

-   <https://ggplot2.tidyverse.org>
-   <https://ggplot2-book.org/>

## Increase Point Size

```{r}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) + geom_point(size = 3)
```

## Add Color

```{r}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) + geom_point(size = 3)
```

## Point Shape by Species

```{r}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species, shape = Species)) + geom_point(size = 3)
```

## The plot from before

```{r}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species, shape = Species)) +
  geom_point() +
  geom_density2d() + # adds a layer with the density lines
  ggtitle('IRIS') +
  theme_light()
```

## Aesthetics

What the different aesthetics do for each kind of variable:

| Aesthetic | Discrete | Continuous |
|------------------------|------------------------|------------------------|
| color | rainbow of colors | gradient |
| size | discrete steps of size | relationship between radius and variable |
| shape | different shape for each | shouldn't and doesn't work |

## Exercise

Make a small subset of the diamonds data:

```{r}
d2 <- sample_n(diamonds,1000)
```

Then generate this plot:

> \*\* removed image so you can knit the file \*\*

```{r, echo = T}

```

## Boxplots!

As always, you can see the help via `?geom_boxplot()` to see your options for arguments.

```{r}
ggplot(mpg, aes(x = class, y = hwy)) +
  geom_boxplot()

# examples of added things: 
# outlier.color = "red"
# notch = TRUE

```

## Histograms

It is common to "save" your work as you go along, building up the plot you want and updating the object as you go along.

```{r}
h <- ggplot(faithful, aes(x = waiting))
h + geom_histogram(binwidth = 5)
```

And then, you can mess around with different things to try them out:

```{r}
# use that same aes setting in h with a new geom for a new type of plot! 
h + geom_freqpoly(binwidth = 5)
```

```{r}
# save as finished product
h <- h + geom_freqpoly(binwidth = 5)
```

## Exercise:

Make a histogram of the carat variable in the diamonds dataset that is available when you load the tidyverse package.

Experiment with different binwidths. What binwidth reveals the most interesting patterns?

There is an additional argument type that is important here - set `binwidth = 1` and look at the far left bin. Add the argument `boundary = 0` to the geom_hist. Why is this better?

```{r}

```

## Titles and Labels

Labels are added with the `labs()` layer. There are many different possibilities, you don't need to use them all every time.

Remember our weird upturned smoother from last class? This example from R4DS uses the subtitle to explain it (and a caption to cite the data).

```{r message=FALSE}
rds <- ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(
    x = "Engine displacement (L)",
    y = "Highway fuel economy (mpg)",
    color = "Car type",
    title = "Fuel efficiency generally decreases with engine size",
    subtitle = "Two seaters (sports cars) are an exception because of their light weight",
    caption = "Data from fueleconomy.gov"
  )

rds
```

## Change It Up

You can change things like color manually:

```{r}
ggplot(faithful, aes(x = waiting)) + geom_histogram(binwidth = 10, fill = "orange", color = "navy")
```

------------------------------------------------------------------------

Or you can change the whole look using a `theme`

```{r}
ggplot(faithful, aes(x = waiting)) + 
  geom_histogram(binwidth = 10) + 
  theme_minimal()
```

------------------------------------------------------------------------

```{r}


rds + theme_dark()
rds + theme_linedraw()
rds + theme_void()

library(ggthemes)
rds + theme_solarized()
rds + theme_economist()
rds + theme_fivethirtyeight()
```

## Exercise:

1.  From the `mpg` data, plot the relationship between `displ` and `hwy`.

-   Create a scatter plot of displ (engine size) vs. hwy (highway mpg) using ggplot().
-   Add geom_point() to display the points.
-   Customize the plot by adding a theme, title and axis labels.

```{r}

```

2.  Modify the scatter plot from Exercise 1 by coloring points based on the class variable.

-   Change the size of the points based on the cyl (number of cylinders) variable.
-   Try using shape instead of size. What happens?

```{r}

```

Before running the command, what will each of these show you? What will they look like?

```{r, eval = F}
ggplot(ggplot2::mpg, aes(cty, hwy, color = factor(cyl))) + geom_jitter()
ggplot(ggplot2::mpg, aes(cty, hwy, color = factor(cyl))) + geom_point()


ggplot(diamonds, aes(carat, price)) + geom_point()

ggplot(economics, aes(date, unemploy)) + geom_line() 

ggplot(mpg, aes(cty)) + geom_histogram()
```

## Bar Graphs

Use `geom_bar()` for a bar chart. You can change the color with either `color =`

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, color = cut))
```

## The better option...

or `fill =`

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut))
```

## Stacked Bars

What happens if we do a fill with a different variable?

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity))
```

## Dodged or Filled Bars

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge")

ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill")
```

## Exercise:

1.  Make a bar plot of species of penguins, where you assign species to the y aesthetic. How is this plot different?

2.  Create a bar chart showing the count of each car class in the mpg dataset.

-   Experiment with fill aesthetics to color bars by drv (drive type: front, rear, 4WD).

```{r}

```

## Bar Graphs for pre-counted data

Suppose you already have the counts:

```{R}
mpg %>% count(class)
```

You would then use `geom_col()` to make this graph, this geom lets you pass an x category and a y count.

```{r}
mpg %>% count(class) %>% ggplot(aes(x = class, y = n)) + geom_col()
```

## Scales

A scatterplot of the gapminder dataset.

General statistics note: When data is "squashed" up against 0 like this, you often want to take a `log()` transform of the data.

```{r}
library(gapminder)
p <- ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) # just initializes
p + geom_point()
```

## Log Transform

We do this with a `scale`!

```{r}
p + geom_point() + scale_x_log10() ## Do we like this? 

# Yes - let's make that stick
p <- p + geom_point() + scale_x_log10()
```

## Add Color

```{r}
#  continent by color: MAP continent variable to aesthetic color
p <- ggplot(gapminder, aes(x = gdpPercap, y = lifeExp, color = continent)) 
p <- p +  geom_point() + scale_x_log10() 
p
```

## Transparency and Size

The previous plot is what we would call *overplotted*... it's busy and you can't really fully see.

We changed the size of the points earlier, but now we're going to change the size and make them more transparent.

```{r}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp, color = continent)) + 
  scale_x_log10() +
  geom_point(alpha = 1/2, size = 2)
```

## Adding Fitted Lines

```{r}
# add a fitted, smoothed, curve 
p + geom_point() + geom_smooth()
```

## Remove the Standard Error Curves

We can remove these intervals by setting the standard error equal to `F`

```{r}
p + geom_point() + geom_smooth(lwd = 2, se = FALSE)
```

Note: these are often a good idea to keep in... they let you know the general "strength" of the model.

## Fit a Linear Model

You can choose which method the smooth uses, in this case we choose "lm", for "linear model" - gives you the linear regression line for these variables.

```{r}
p + geom_point() + geom_smooth(lwd = 3, se = FALSE, method = "lm")
```

## Seeing the Different Continents

If we're interested in which continent has what trend, we can't see that super well at this point.

So, we'll split them up with something called `facet`s.

```{r, echo = F}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp, color = continent)) + 
  scale_x_log10()  + geom_point(alpha = 0.5) +
  geom_smooth(lwd = 1, se = FALSE) +
  facet_wrap(~ continent)
```

## Facetting

Facetting is a way you can put those factors to work for you.

`facet_wrap` splits up the plots by that factor, giving the plots separately, along with the legend.

```{r}
p + geom_point() +
  facet_wrap(~ continent)
```

```{r}
p + geom_point(alpha = (1/3), size = 3) +
  facet_wrap(~ continent) +
  geom_smooth(lwd = 2, se = FALSE)
```

## Facet Grid

Facetting has two options:

The one we've been using, 'facet_wrap()\` which makes a 1d ribbon wrapped into 2d.

OR you can make a controlled grid of your data using two variables.

`facet_grid()`: 2d grid, `rows ~ cols`,`.` for no split

```{r}
ggplot(mpg, aes(cty, hwy)) + geom_point() + facet_grid(drv ~ cyl )
```

## Exercises

For the gapminder data:

-   Plot lifeExp against year\
-   make mini-plots, split out by continent HINT: use facet_wrap()
-   change the theme to black and white, `theme_bw()`
-   add a fitted smooth line, w/ or w/o facetting\
-   use `dplyr::filter()` to plot lifeExp against year for just one country or continent

```{r}

# Plot lifeExp against year
gapminder %>% ggplot(aes(x = year,y = lifeExp)) + geom_point()


# make mini-plots, split out by continent HINT: use facet_wrap()

gapminder %>% ggplot(aes(x = year,y = lifeExp)) + geom_point() + facet_wrap(~ continent)

# change the theme to black and white, `theme_bw()`

gapminder %>% ggplot(aes(x = year,y = lifeExp)) + geom_point() + facet_wrap(~ continent) + theme_bw()

# add a fitted smooth line, w/ or w/o facetting  

gapminder %>% ggplot(aes(x = year,y = lifeExp)) + geom_point() + theme_bw() + geom_smooth()

# use `dplyr::filter()` to plot lifeExp against year for just one country or continent

gapminder %>% filter( continet == "Europe" ) %>% ggplot(aes(x = factor (year) ,y = lifeExp)) + geom_point() + theme_bw() + geom_smooth()


```

## Saving Our Work

In R, you can always do a manual click-save to save/export plots, **but** that is usually a pain, and never reproducible. - Suppose you need to make a plot for every state. Are you going to go in and click for each state? No.

```{r, eval = F}
print(p) ## prints to the "plots" output
ggsave("your_file_name.pdf") ## saves the current plot output to your current working directory
ggsave("your_file_name.pdf",plot = p) ## saves the plot of your object p to your current working directory
```

## Line Plots

How can we "connect the dots" for one country? i.e. make a spaghetti plot?

```{r}
#' plot lifeExp against year
y <- ggplot(gapminder, aes(x = year, y = lifeExp))

y + facet_wrap(~ continent) + geom_line() # uh, no
```

## Much Better

Think through what you want to do. We want the lines grouped by country, so we find the command that does that.

```{r}
y + facet_wrap(~ continent) + geom_line(aes(group = country)) # yes!
```

## Subsetting

Sadly, ggplot() does not have a 'subset =' argument. So, do that 'on the fly' with `subset(..., subset = ...)` or by piping to filter

```{r, eval = F}
ggplot(subset(gapminder, country == "Canada"),
       aes(x = year, y = lifeExp)) + geom_line() + geom_point()
```

```{r}
ggplot(gapminder %>% filter(country == "Canada"),
       aes(x = year, y = lifeExp)) + geom_line() + geom_point()
```

## Combos with `dplyr`

We can just look at four countries:

```{r}
jCountries <- c("Canada", "Denmark", "Poland", "Mexico")
ggplot(gapminder::gapminder %>% filter(country %in% jCountries),
       aes(x = year, y = lifeExp, linetype = country)) + geom_line() + geom_point()
```

## All Sorts of Options and Extras

Remember this?

```{r}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) +  scale_x_log10()  + geom_point()
```

------------------------------------------------------------------------

Let's make it a heatmap!

```{r}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) +
  scale_x_log10() + geom_bin2d()
```

## Summary

The challenge of visualization, like any other design problem, is to identify an optimal solution with your constraints.

Consider:

-   The question you're answering
-   The type of data you have
-   The limitations of the human visual processing system (aka DO NOT USE PIE CHARTS)
-   The spatial limitations of your medium
    -   Pixels on the screen, inches on the page, dimension of your medium.

## Resources

-   The documentation's reference: <https://ggplot2.tidyverse.org/reference>
-   A "cookbook" with "recipes" for plots. <http://www.cookbook-r.com/Graphs/index.html>
-   R for Data Science - has data visualization chapters. <https://r4ds.hadley.nz>
-   And, finally, the cheatsheet. <http://bit.ly/ggplot2-cheatsheet>
